---
title: "CS109b Homework 2 Submission"
author: "Ihsaan Patel"
date: "February 11, 2017"
output: pdf_document
---

# Helper Functions

```{r helper_functions}
library(ggplot2)
library(gam)
library(boot)

# Function to compute k-fold cross-validation accuracy for a given classification model
cv_accuracy = function(data, param_val, k) {
  # Input: 
  #   'model' - a fitted classification model
  #   'data' - data frame with training data set used to fit the model
  #   'k' - number of folds for CV
  # Output:
  #   'cv_accuracy' - cross-validation accuracy for the model
  
  num_param = length(param_val) # Number of parameters
  cv_acc = rep(0., num_param)
  
  # Iterate over parameter values
  for(i in 1:num_param){
    gam_formula <- as.formula(sprintf("HeartDisease ~ s(Age, spar=%1$f) + Sex + s(RestBP,spar=%1$f) + ExAng + ChestPain + Thal", param_val[i]))
    model <- gam(gam_formula, family = binomial(link = "logit"), data = data)
    acc <- 1 - cv.glm(data, model, K = k)$delta[1]
    cv_acc[i] <- acc
  }
  return(cv_acc)
}

```

# Problem 1: Heart Disease Diagnosis

## Visual Inspection of Data

```{r load_data}
# Load data
train_dataset1 <- read.table("dataset_1_train.txt",header = TRUE, sep = ",")
test_dataset1 <- read.table("dataset_1_test.txt",header = TRUE, sep = ",")

# Plot and visualize Age
ggplot(train_dataset1, mapping = aes(x = Age , fill = HeartDisease)) + geom_histogram(binwidth=1, alpha=.5, position="identity") + ggtitle("Age & Heart Disease")
```

Older people appear to have much higher incidents of heart disease, particularly between the ages of 55 and 70.

```{r plot_sex}
# Plot and visualize Sex
ggplot(train_dataset1, mapping = aes(x = Sex, fill = HeartDisease)) + geom_bar(position = "dodge") + ggtitle("Sex & Heart Disease")
```

There appears to be a much higher incidence of heart disease in one of the sexes

```{r plot_chest_pain}
# Plot and visualize Chest Pain
ggplot(train_dataset1, mapping = aes(x = ChestPain, fill = HeartDisease)) + geom_bar(position = "dodge") + ggtitle("Sex & Heart Disease")
```

There's a much higher incidence of heart disease in individuals with asymptomatic chest pain.

```{r plot_rest_bp}
# Plot and visualize RestBP
ggplot(train_dataset1, mapping = aes(x = RestBP, fill = HeartDisease)) + geom_histogram( alpha=.5, position="identity") + ggtitle("RestBP & Heart Disease")
```
Higher values of resting BP appear to be associated with heart disease

```{r plot_ex_ang}
# Plot and visualize ExAng
ggplot(train_dataset1, mapping = aes(x = ExAng, fill = HeartDisease)) + geom_bar(position = "dodge") + ggtitle("ExAng & Heart Disease")
```

People who have ExAng appeart to have a much higher incidence of heart disease

```{r plot_thal}
# Plot and visualize Thal
ggplot(train_dataset1, mapping = aes(x = Thal, fill = HeartDisease)) + geom_bar(position = "dodge") + ggtitle("Thal & Heart Disease")
```

People  with both fixed and particularly reversable Thal have much incidences of heart disease

## GAM Model

```{r cv_gam_model, warning=FALSE}
# Tune smoothing parameter through k-fold cross validation
spar_to_validate <- c(1.:100.)/100.
gam_cv_scores <- cv_accuracy(train_dataset1, spar_to_validate, 5)
spar_gam_max_parameter <- spar_to_validate[which.max(gam_cv_scores)]
ggplot(data = data.frame(parameter = spar_to_validate, parameter_score = gam_cv_scores), aes(x = parameter, y = parameter_score, group = 1)) + geom_line() + geom_point() + ggtitle(sprintf("Smoothing Cross Validation | Max Parameter: %.2f", spar_gam_max_parameter))

```

```{r fit_gam_model, warning=False}
gam_formula <- as.formula(sprintf("HeartDisease ~ s(Age, spar=%1$f) + Sex + s(RestBP,spar=%1$f) + ExAng + ChestPain + Thal", spar_gam_max_parameter))

model.gams <- gam(gam_formula, family = binomial(link = "logit"), data = train_dataset1)

pred.gams <- predict(model.gams, newdata = test_dataset1, type = "response")
cm.gams <- as.matrix(table(Actual = test_dataset1$HeartDisease, Predicted = pred.gams>.5))
accuracy.gams <- sum(diag(cm.gams))/ sum(cm.gams)

print(sprintf("GAM Model Classification Accuracy: %.4f", accuracy.gams))
```
* The smoothing spline basis cannot be applied to categorical predictors because the weighted average would produce either the predictors actual value or some meaningless value between predictors. For instance, if 0 represents male and 1 represents female, and a weighted average produced .8, this would not be useful as it doesn't correspond to either male or female.
* R appears to one-hot encode categorical predictors automatically whereas this needed to be done manually in sklearn in Python

## Predictor Smoothing

```{r plot_gam_smooth}
plot.gam(model.gams, se = TRUE, terms = c(sprintf("s(Age, spar = %.2f)",spar_gam_max_parameter), sprintf("s(RestBP, spar = %.2f)",spar_gam_max_parameter)))
```

There doesn't appear to be any benefit smoothing the splines due to the large standard errors for non-zero coefficient values.

## Likelihood Tests

### (i) GAM with only Intercept

```{r gam_with_intercept}
model.gami <- gam(HeartDisease ~ 1, family = binomial(link = "logit"), data = train_dataset1)
anova(model.gams, model.gami, test = "Chi")
```

The gam model that includes all predictors with smoothing on the continuous ones performs better than the gam model with only the intercept at a significance level of 0.001.

### (ii) GAM with Categorical

```{r gam_with_intercept}
model.gamcat <- gam(HeartDisease ~ Sex + ChestPain + ExAng + Thal, family = binomial(link = "logit"), data = train_dataset1)
anova(model.gams, model.gamcat, test = "Chi")
```

The gam model that includes all predictors with smoothing on the continuous ones performs better than the gam model with only categorical predictors at a significance level of 0.05.

### (iii) GAM with Linear Predictors

```{r gam_with_intercept}
model.gamlin <- gam(HeartDisease ~ Age + Sex + ChestPain + RestBP + ExAng + Thal, family = binomial(link = "logit"), data = train_dataset1)
anova(model.gams, model.gamlin, test = "Chi")
```

The gam model that includes all predictors with smoothing on the continuous ones performs better than the gam model with all predictors entered linearly, however the improvement is not statistically significant.
